Procurements of the Canadian Government, Runaway Military Contracts

### 2024-11-29 paper/paper.qmd
TODO Refs
added top-buyers chart

TODO Talk about top 5 buyers and other
```{r fig-top-buyers}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: template
#| fig-align: center

#TODO color each buyer with the correct color

@@

# Create a new column 'buyer_group' where buyers not in top 5 are labeled as 'Other'
buyer_totals_grouped <- buyer_totals %>%
  mutate(buyer_group = ifelse(buyer %in% top_five_buyers, buyer, "Other")) %>%
  group_by(buyer_group) %>%
  summarise(total_amount = sum(total_amount, na.rm = TRUE)) %>%
  arrange(desc(total_amount))

# Plot the bar chart
ggplot(buyer_totals_grouped, aes(x = reorder(buyer_group, total_amount), y = total_amount)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  labs(title = "Top 5 Buyers by Total Amount Spent and Others",
       x = "Buyer",
       y = "Total Amount Spent (CAD)") +
  scale_y_continuous(labels = comma) +
  theme_minimal()

@@

TODO add spending by selected buyers graph

@@

TODO Ref RSelenuim

>>>>

TODO Talk about top 5 buyers and other

```{r fig-pie-buyers}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Top 5 Buyers by Total Amount Spent and Others (Pie Chart)
#| fig-align: center

@@

# Create a new column 'buyer_group' where buyers not in top 5 are labeled as 'Other'
buyer_totals_grouped <- buyer_totals %>%
  mutate(buyer_group = ifelse(buyer %in% top_five_buyers, buyer, "Other")) %>%
  group_by(buyer_group) %>%
  summarise(total_amount = sum(total_amount, na.rm = TRUE)) %>%
  arrange(desc(total_amount)) %>%
  mutate(percentage = total_amount / sum(total_amount) * 100)

# TODO: Assign specific colors to each buyer group
# Example color assignments (replace with desired colors)
buyer_colors <- c(
  "National Defence" = "#1f77b4",
  "Public Services and Procurement Canada" = "#ff7f0e",
  "Shared Services Canada" = "#2ca02c",
  "Public Health Agency of Canada" = "#d62728",
  "Fisheries and Oceans Canada" = "#9467bd",
  "Other"  = "#8c564b"
)

# Plot the pie chart
ggplot(buyer_totals_grouped, aes(x = "", y = total_amount, fill = buyer_group)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar(theta = "y") +
  labs(
    title = "Top 5 Buyers by Total Amount Spent and Others",
    fill = "Buyer"
  ) +
  scale_fill_manual(values = buyer_colors) +
  theme_void() + # Removes background, grid, and numeric labels
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold")
  ) +
  geom_text(aes(label = paste0(round(percentage, 1), "%")),
            position = position_stack(vjust = 0.5),
            color = "white",
            size = 4)
```

@@

TODO add spending by selected buyers graph
```{r fig-top-buyers}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: template
#| fig-align: center

# Filter out records from 2019
df_filtered <- df %>%
  filter(format(award_date, "%Y") != "2019")

# Extract the year from award_date
df_filtered$year <- format(df_filtered$award_date, "%Y")

# Define the selected buyers
selected_buyers <- c(
  "National Defence",
  "Public Services and Procurement Canada",
  "Shared Services Canada",
  "Public Health Agency of Canada",
  "Fisheries and Oceans Canada"
)

# Add a new column for buyer categories
df_filtered <- df_filtered %>%
  mutate(buyer_category = ifelse(buyer %in% selected_buyers, buyer, "Other"))

# Summarize total spending per year for each buyer category
grouped_df <- df_filtered %>%
  group_by(year, buyer_category) %>%
  summarize(total_spending = sum(amount, na.rm = TRUE)) %>%
  ungroup()

# Ensure the order of the groups matches the specified order, with "Other" last
grouped_df$buyer_category <- factor(
  grouped_df$buyer_category,
  levels = c(selected_buyers, "Other")
)

# Scale the total_spending to billions
grouped_df <- grouped_df %>%
  mutate(total_spending_billion = total_spending / 1e9)

# Create the grouped bar chart
ggplot(data = grouped_df, aes(x = year, y = total_spending_billion, fill = buyer_category)) +
  geom_bar(
    stat = "identity",
    position = position_dodge(width = 0.9), # Adjust dodge width
    width = 0.8 # Adjust bar width
  ) +
  scale_fill_manual(
    values = c(
      "National Defence" = "steelblue",
      "Public Services and Procurement Canada" = "orange",
      "Shared Services Canada" = "green",
      "Public Health Agency of Canada" = "purple",
      "Fisheries and Oceans Canada" = "red",
      "Other" = "gray"
    )
  ) +
  scale_y_continuous(
    labels = comma_format(accuracy = 1), # Formats y-axis labels without decimal places
    breaks = pretty_breaks(n = 5), # Ensures approximately 5 breaks like 5, 10, 15
    expand = expansion(mult = c(0, 0.05)) # Adds a small space above the highest bar
  ) +
  labs(
    title = "Spending by Selected and Other Buyers Per Year",
    subtitle = "Amounts are in Billions of CAD",
    x = "Year",
    y = "Total Spending (Billions)",
    fill = "Buyer",
    caption = "."
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1), # Rotate x-axis labels for readability
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12),
    plot.caption = element_text(hjust = 0.5, size = 10, face = "italic") # Centered caption
  )

@@

TODO Ref:

RSelenuim
VGAM
stringr
dplyr
tidyverse
zoo
kableExtra
knitr
scales
lubridate
ggplot2
rvest
readr
testthat
